---
title: "Exercise 2C: Local Measures of Spatial Autocorrelation"
author: "Oh Jia Wen"
date: 11/24/2023
date-modified: "last-modified"
execute: 
  echo: true
  eval: true
  warning: false
---

# 1. Getting Started

## 1.1 Install and launching R packages

The code chunk below uses **p_load()** of pacman package to check if `sf`, `spdep`, `tmap`, and `tidyverse` packages are installed into the R environment. If they are, then they will be launched into R.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

## 1.2 Importing the OD data

We will be using two data sets for this exercise. Data were retrieved on 19th Nov 2023. They are :

1.  **Hunan country boundary layer\***. *(data is in ESRI shapefile format) - Geospatial data*
2.  **Hunan_2012.csv\*.** *(data is in csv file) - Attribute table*

In this exercise, we are interested to examine the spatial pattern of GDPPC (a.k.a GPD per Capital) of Hunan Provice, People Republic of China.

### 1.2.1 Importing Hunan country boundary layer

The code chunk below uses `st_read()` of **sf** package to import the [1st data set]{.underline} into R. The imported shapefile will be simple features object of sf.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### 1.2.2 Importing Hunan_2012.csv

Next, we will import the [2nd dataset (csv)]{.underline} into R. We will use `read_csv()` of **readr** package. The output is in R dataframe class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## 1.3 Performing relational join

After importing, we will update the attribute table of **hunan**'s Spatial Polygons Data Frame with the attribute fields of **hunan2012** dataframe. We will performed a `left_join()` with the aid of **dplyr** package.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4,7,15)
```

We will be joining both tables by County. By doing the left_join, we will combined the 8 variables from **hunan**, with 29 variables from **hunan2012** and uses `select()` to filter for the variables that we are interested in.

## 1.4 . Visualising Regional Development Indicator

After joining, we will do a quick visualization. We will be using the `qtm()` of **tmap** package to prepare a basemap and a choropleth map to see the distribution of ***GDPPC 2012***.

```{r}
#| code-fold: true
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

# 2. Cluster and Outlier Analysis

In Exercise 2B, we covered the Global Indicators of Spatial Association. 

```{r}
#| code-fold: true
# incorporate global SA to do local 

#computing contiguity spatial weights
wm_q <- poly2nb(hunan, 
                queen=TRUE)

#row-standardized weights matrix
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
```

In this exercise, we will be covering the Local Indicators of Spatial Association or LISA to evaluate existence of clusters in the spatial arrangement of a given variable.

We will be applying local Moran' I to detect cluster and/or outliers from GDP per capita 2012 of Hunan Province, PRC.

::: panel-tabset
# Local Moran's I Test

We will perform Local Moran's I statistical testing using `localmoran()` of **spdep**. Given a set of Z.Li values, it computes **Ii** , and a listw object providding neighbor weighting information for the polygon associated with the Z.Ii values.

The code chunks below are used to compute local Moran's I of GDPPC2012 at the county level.

```{r}
#| code-fold: true
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)
```

::: {.callout-note title="Note about localmoran() " collapse="true"}
## Note about `localmoran()`

`localmoran()` returns a matrix of values whose columns are:

1.  "Ii" : the local Moran's I statistics

2.  "E.Ii" : the **expectation** of local moran statistics under the randomisation hypothesis

3.  "Var.Ii" : the **variance** of local moran statistic under the randomisation hypothesis

4.  "Z.Ii" : the **standard deviation** of local moran statistic

5.  "Pr()" : the **p-value** of local moran statistic
:::

To see the content of the local Moran matrix, we can use printCoefmat() :

```{r}
#| code-fold: true
printCoefmat(data.frame(
  localMI[fips,], 
  row.names=hunan$County[fips]),
  check.names=FALSE)
```

# Mapping

Notably, we should append the local Moran's I dataframe (i.e. localMI) onto hunan Spatial Polygon Data Frame before mapping the local Moran's I map. To perform the tasks, we can use the code below:

## Step 1: Append Local Moran's I with SPDF

```{r}
hunan.localMI <- cbind(hunan,localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

## Step 2: Map local Moran's I values

Using choropleth mapping function of **tmap** package, we can plot the local Moran's I values:

```{r}
#| code-fold: true
tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty",
          palette = "RdBu",
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)
```

## Step 3: Map local Moran's I p-values

As seen above, Ii contains both positive and negative values. Thus, it is useful for us to consider the p-values for each values.

```{r}
#| code-fold: true
tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)
```

## Step 4 (Optional) : Map both local Moran's I values and p-values

For effective interpretation, we can plot both I-values and corresponding p-values next to one another.

```{r}
#| code-fold: true
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)

```
:::

# 3. Creating LISA Cluster Map

We are able to observe significant locations color coded by type of spatial autocorrelation. To begin with, we will plot the Moran Scatterplot

::: panel-tabset
# 1) Plot Moran Scatterplot

Firstly, we will plot the Moran Scatterplot of GDPPC 2012 by using `moran.plot()` of **spdep**:

```{r}
#| code-fold: true
nci <- moran.plot(hunan$GDPPC, rswm_q,
           labels=as.character(hunan$County), 
                  xlab="GDPPC 2012", 
                  ylab="Spatially Lag GDPPC 2012")
```

From the result above, the plot is split into 4 quadrants. Quadrant 2 (top right) are the **high-high locations**. They belongs to area with high GDPPC and are surrounded by other areas with average level of GDPPC.

# 2) Add Standardised Variable

Next, we will use `scale()` to centers and scale the variable. **Centering** is done by subtracting the mean (omitting NAs) for the corresponding columns, and **Scaling** is done by dividing the (centered) varialbe by their standard deviations.

At the end, `as.vector()` is added to ensure data type is a vector.

```{r}
#| code-fold: true
#add scale to the previous step 
hunan$Z.GDPPC <- scale(hunan$GDPPC) %>% 
  as.vector 

#plot moran scatterplot
nci2 <- moran.plot(hunan$Z.GDPPC, rswm_q,
                   labels=as.character(hunan$County),
                   xlab="z-GDPPC 2012", 
                   ylab="Spatially Lag z-GDPPC 2012")
```

# 3) Prep LISA Map classes

The code chunks below show the steps to prepare a LISA cluster map.

## Step 1: Vectorize

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
```

## Step 2: Derive Spatially lagged variable of interest and centers it around its mean

```{r}
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)     
```

## Step 3: Center the local Moran's around the mean

```{r}
LM_I <- localMI[,1] - mean(localMI[,1])    
```

## Step 4: Set statistical significance level for the local Moran

```{r}
signif <- 0.05       
```

## Step 5: Define the quadrants
1 = low-low
2 = low - high
3 = high - low
4 = high - high

```{r}
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4      
```

## Step 6: Place non-significant Moran in Category 0 

```{r}
quadrant[localMI[,5]>signif] <- 0
```

## Final Output: 

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)     
LM_I <- localMI[,1]   
signif <- 0.05       
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4    
quadrant[localMI[,5]>signif] <- 0
```

# 4) Plotting LISA map
With the preparation, we can now plot the LISA map by plotting both local Moran's I values map and its corresponding p-values map next to one another. 

```{r}
#| code-fold: true

gdppc <- qtm(hunan, "GDPPC")

hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap <- tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

#to plot I-value and P-value map
tmap_arrange(gdppc, LISAmap, 
             asp=1, ncol=2)
```
:::

# 4. Hot Spot and Cold Spot Area Analysis


