---
title: "In-class Exercise 1"
author: "Oh Jia Wen"
date: 11/18/2023
date-modified: "last-modified"
execute: 
  echo: true
  eval: true
  warning: false
---

# 1. Getting Started

## 1.1 Install and launching R packages

The code chunk below load the following packages:

-   **tmap** : for thematic mapping

-   **sf** : for geospatial data handling

-   **tidyverse**: for non-spatial data handling

```{r}
pacman::p_load(tmap, sf, tidyverse)
```

## 1.2 Importing the OD data

Firstly, we will import the [Passenger Volume by Origin Destination Bus Stops]{.underline} data set downloaded from LTA DataMall by using `read_csv()` of **readr** package.

```{r}
#| eval: false
odbus <- read_csv("data/origin_destination_bus_202308.csv")
```

A quick check of odbus tibble data frame shows that the value in ORIGIN_PT_CODE.

```{r}
#| eval: false
odbus$ORIGIN_PT_CODE <- as.factor(odbus$ORIGIN_PT_CODE)

odbus$DESTINATION_PT_CODE <- as.factor(odbus$DESTINATION_PT_CODE)
```

### 1.3 Extracting the study data

For the purpose of this exercise, we will extract commuting flows on weekday between 7 to 9 o'clock.

```{r}
#| eval: false
origtrip_7_9 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 7 &
           TIME_PER_HOUR <= 9 ) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
  ungroup()
```

**Two geospatial data** will be used in this exercise, they are:

```{r}
#| eval: false

busstop <- st_read(dsn = 'data', layer = 'BusStop') %>%
  st_transform(crs= 3414)

```

```{r}
mpsz <- st_read(dsn = 'data', layer = 'MPSZ-2019') %>%
  st_transform(crs= 3414)

mpsz
```
